### 2017-07-07 paper/extensions.tex

To transfer this mathematical result to probabilistic programs, we want to 
impose a restriction on LexRSM maps that ensures that all components of a 
LexRSM map have, from each reachable configuration, an expected one-step 
increase of at most $c$. Here $c$ can be a constant, but it can also be a value 
that depends on the initial configurations: this is to handle cases where some 
variables are periodically reset to a value related to the initial variable 
values, such as variable $y$ in Figure~\ref{fig:uniint4}. To this end, let 
$\program$ be a \PP{} with a pCFG $\pCFG_\program$ and let 
$\vec{\lem}=(\lem_1,\dots,\lem_n)$ be an 
$n$-dimensional $1$-LexRSM map for $\program$. Consider an $n$-dimensional 
vector 
$\vec{\bar{c}}=(\bar{c}_1,\dots,\bar{c}_n)$ whose each component is an 
expression over variables of 
$\program$. We say 
that $\vec{\lem}$ has 
$\vec{\bar{c}}$-bounded ECI w.r.t. invariant map $\inv$ if the following holds 
for each initial configuration $(\locinit,\vecinit)$ with $\vecinit\in 
\vecinitset$: for 
each 
configuration $(\loc,\vec{x})$ with $\vec{x}\in \inv(\loc)$ and generalized 
transition $\tilde{\tau}$ 
of 
$\pCFG_\program$ outgoing from $(\loc,\vec{x})$ it holds that if $j$ is 
the smallest index such that 
$\tilde{\tau}$ is $1$-ranked by $\lem_j$ from $(\loc,\vec{x})$, then for all 
$j'>j$ the gen. transition $\tilde{\tau}$ is $f$-ranked by $\lem_{j'}$ from 
$(\loc,\vec{x})$, where $f=-c_{j'}(\vecinit)$. From 
Theorem~\ref{thm:runtime-bound} we have the following:

\begin{corollary}
\label{col:runtime-progs}
Let $\program$ be a probabilistic program. Assume that there exists an 
$n$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_d)$ for 
$\program$ supported 
by some 
invariant map $\inv$, scuh that $\vec{\lem}$ has $\vec{\bar{c}}$-bounded ECI 
(w.r.t. $\inv$) 
for some vector of expressions $\vec{\bar{c}}=(\bar{c}_1,\dots,\bar{c}_n)$. 
Then under each scheduler $\sigma$ and for each initial valuation of program 
variables $\vecinit\in\vecinitset$ it holds $\E^{\sigma}_{\vecinit}[\ttime]\leq
\sum_{j=1}^{n}\lem_j(\locinit,\vecinit)\cdot(1+\sum_{j'= j+1}^{n}\prod_{j'}^{n} 
\bar{c}_{j'}(\vecinit)).$
%\sum_{j=1}^{n}\lem_j(\locinit,\vecinit)\cdot(\bar{c}_j(\vecinit))^{n-j}$.
\end{corollary}

\begin{example}
Consider the program in Figure~\ref{fig:uniint4}. For the program we have an 
invariant map $\inv$ with $\inv(\loc_0)=x\geq -1$, $\inv(\loc_1)=\inv(\loc_4) = 
x\geq 0 
\wedge z\geq 0$, $\inv(\loc_2)=\inv(\loc_3)= x\geq 0 \wedge y \geq 0
\wedge z\geq 0$, and $\inv(\loc^{\lout})=\mathit{true}$. Further, there exists 
the following 
$3$-dimensional 
$1$-LinLexRSM map $\vec{\lem}$ supported by $\inv$: 
$\vec{\lem}(\loc_0)=(2x+3,0,2)$, $\vec{\lem}(\loc_1)=(2x+3,0,1)$, 
$\vec{\lem}(\loc_2)=(2x+2,y,1)$, $\vec{\lem}(\loc_3)=(2x+2,y,0)$, 
$\vec{\lem}(\loc_4)=(2x+2,0,0)$, and $\vec{\lem}(\loc^{\lout})=(0,0,0)$. It is 
easy to check that $\vec{\lem}$ is has $(0,z,2)$-bounded ECI. Applying 
Corollary~\ref{col:runtime-progs} yields $\E[\ttime]\leq (2x_{\mathit{init}} + 
3)\cdot(1+3z_{\mathit{init}})+0+2= 6x_{\mathit{init}}z_{\mathit{init}} + 
2x_{\mathit{init}} + 9z_{\mathit{init}}+5$. Note that this quadratic bound is 
asymptotically optimal for the given program.
\end{example}

Now assume that we have synthesized a $1$-LexRSM map $\vec{\lem}$ and we want 
to check if there exists $\bar{\vec{c}}$ such that $\vec{\lem}$ has 
$\bar{\vec{c}}$-bounded ECI. In the linear setting (i.e. the program is an 
\APP{}, masp $\lem$ and $\inv$ are linear, and we seek $\vec{\bar{c}}$ which is 
a vector of affine expressions) we can encode the existence of $\vec{\bar{c}}$ 
into a system of linear inequalities in the similar way as the existence of a 
linear LexRSM maps was encoded in Section~\ref{sec:lex-programs}. That is, we 
set up a linear template with unknown coefficients for each component of 
$\vec{\bar{c}}$ and using Farkas's lemma we set up a system of linear 
constraints, which includes the unknown coefficients as variables, encoding the 
fact that $\vec{\lem}$ has $\bar{\vec{c}}$-bounded ECI. Details are provided 
in~\AppendixMaterial. In this way, we can check in polynomial time if 
Corollary~\ref{col:runtime-progs} can be applied to $\vec{\lem}$, and if yes, 
we can synthesize the witness vector $\bar{\vec{c}}$. Since $\bar{\vec{c}}$ 
consists of affine expressions, Corollary~\ref{col:runtime-progs} provides a 
polynomial (in the size of 
initial variable valuation) upper bound on expected runtime.

>>>>

To transfer this mathematical result to probabilistic programs, we want to 
impose a restriction on LexRSM maps that ensures that all components of a 
LexRSM map have, from each reachable configuration, an expected one-step 
increase of at most $c$. Here $c$ can be a constant, but it can also be a value 
that depends on the initial configurations: this is to handle cases where some 
variables are periodically reset to a value related to the initial variable 
values, such as variable $y$ in Figure~\ref{fig:uniint4}. To this end, let 
$\program$ be a \PP{} with a pCFG $\pCFG_\program$ and let 
$\overrightarrow{\lem}=(\lem_1,\dots,\lem_n)$ be an 
$n$-dimensional $1$-LexRSM map for $\program$. Consider an $n$-dimensional 
vector 
$\vec{\bar{c}}=(\bar{c}_1,\dots,\bar{c}_n)$ whose each component is an 
expression over variables of 
$\program$. We say 
that $\overrightarrow{\lem}$ has 
$\vec{\bar{c}}$-bounded ECI w.r.t. invariant map $\inv$ if the following holds 
for each initial configuration $(\locinit,\vecinit)$ with $\vecinit\in 
\vecinitset$: for 
each 
configuration $(\loc,\vec{x})$ with $\vec{x}\in \inv(\loc)$ and generalized 
transition $\tilde{\tau}$ 
of 
$\pCFG_\program$ outgoing from $(\loc,\vec{x})$ it holds that if $j$ is 
the smallest index such that 
$\tilde{\tau}$ is $1$-ranked by $\lem_j$ from $(\loc,\vec{x})$, then for all 
$j'>j$ the gen. transition $\tilde{\tau}$ is $f$-ranked by $\lem_{j'}$ from 
$(\loc,\vec{x})$, where $f=-c_{j'}(\vecinit)$. From 
Theorem~\ref{thm:runtime-bound} we have the following:

\begin{corollary}
\label{col:runtime-progs}
Let $\program$ be a probabilistic program. Assume that there exists an 
$n$-dimensional $\eps$-LexRSM map $\overrightarrow{\lem}=(\lem_1,\dots,\lem_d)$ 
for 
$\program$ supported 
by some 
invariant map $\inv$, such that $\overrightarrow{\lem}$ has 
$\vec{\bar{c}}$-bounded ECI 
(w.r.t. $\inv$) 
for some vector of expressions $\vec{\bar{c}}=(\bar{c}_1,\dots,\bar{c}_n)$. 
Then under each scheduler $\sigma$ and for each initial valuation of program 
variables $\vecinit\in\vecinitset$ it holds $\E^{\sigma}_{\vecinit}[\ttime]\leq
\sum_{j=1}^{n}\lem_j(\locinit,\vecinit)\cdot(1+\sum_{j'= j+1}^{n}\prod_{j'}^{n} 
\bar{c}_{j'}(\vecinit)).$
%\sum_{j=1}^{n}\lem_j(\locinit,\vecinit)\cdot(\bar{c}_j(\vecinit))^{n-j}$.
\end{corollary}

\begin{example}
Consider the program in Figure~\ref{fig:uniint4}. For the program we have an 
invariant map $\inv$ with $\inv(\loc_0)=x\geq -1$, $\inv(\loc_1)=\inv(\loc_4) = 
x\geq 0 
\wedge z\geq 0$, $\inv(\loc_2)=\inv(\loc_3)= x\geq 0 \wedge y \geq 0
\wedge z\geq 0$, and $\inv(\loc^{\lout})=\mathit{true}$. Further, there exists 
the following 
$3$-dimensional 
$1$-LinLexRSM map $\overrightarrow{\lem}$ supported by $\inv$: 
$\overrightarrow{\lem}(\loc_0)=(2x+3,0,2)$, 
$\overrightarrow{\lem}(\loc_1)=(2x+3,0,1)$, 
$\overrightarrow{\lem}(\loc_2)=(2x+2,y,1)$, 
$\overrightarrow{\lem}(\loc_3)=(2x+2,y,0)$, 
$\overrightarrow{\lem}(\loc_4)=(2x+2,0,0)$, and 
$\overrightarrow{\lem}(\loc^{\lout})=(0,0,0)$. It is 
easy to check that $\overrightarrow{\lem}$ is has $(0,z,2)$-bounded ECI. 
Applying 
Corollary~\ref{col:runtime-progs} yields $\E[\ttime]\leq (2x_{\mathit{init}} + 
3)\cdot(1+3z_{\mathit{init}})+0+2= 6x_{\mathit{init}}z_{\mathit{init}} + 
2x_{\mathit{init}} + 9z_{\mathit{init}}+5$. Note that this quadratic bound is 
asymptotically optimal for the given program.
\end{example}

Now assume that we have synthesized a $1$-LexRSM map $\overrightarrow{\lem}$ 
and we want 
to check if there exists $\bar{\vec{c}}$ such that $\overrightarrow{\lem}$ has 
$\bar{\vec{c}}$-bounded ECI. In the linear setting (i.e. the program is an 
\APP{}, masp $\lem$ and $\inv$ are linear, and we seek $\vec{\bar{c}}$ which is 
a vector of affine expressions) we can encode the existence of $\vec{\bar{c}}$ 
into a system of linear inequalities in the similar way as the existence of a 
linear LexRSM maps was encoded in Section~\ref{sec:lex-programs}. That is, we 
set up a linear template with unknown coefficients for each component of 
$\vec{\bar{c}}$ and using Farkas's lemma we set up a system of linear 
constraints, which includes the unknown coefficients as variables, encoding the 
fact that $\overrightarrow{\lem}$ has $\bar{\vec{c}}$-bounded ECI. Details are 
provided 
in~\AppendixMaterial. In this way, we can check in polynomial time if 
Corollary~\ref{col:runtime-progs} can be applied to $\overrightarrow{\lem}$, 
and if yes, 
we can synthesize the witness vector $\bar{\vec{c}}$. Since $\bar{\vec{c}}$ 
consists of affine expressions, Corollary~\ref{col:runtime-progs} provides a 
polynomial (in the size of 
initial variable valuation) upper bound on expected runtime.

### 2017-07-07 paper/lexicographic-programs.tex

\begin{definition}[Measurable Maps and Linear Expression Maps]
A 1-dimensional \emph{measurable map} for a \PP{} $\program$ is a  
real-valued function $\lem$ 
assigning to each program location $\loc$ of $\pCFG_P$ a Borel-measurable function $\lem(\loc)$  of program variables, i.e. each $\lem(\loc)$  is a function of type $\Rset^{|\pvars|}\rightarrow \Rset$. As a special case, if all the functions $\lem(\loc)$ are affine, then we call $\lem$ a 1-dimensional \emph{linear expression map (LEM)}. 
Am $n$-dimensional measurable/linear expression map is a vector $\vec{\lem}=(\lem_1,\dots,\lem_n)$ of 1-dimensional measurable/linear expression maps. 
\end{definition}

@@

\begin{definition}[Lexicographic Ranking Supermartingale Map]
Let $\eps>0$. An $n$-dimensional \emph{lexicographic $\eps$-ranking supermartingale map} ($\eps$-LexRSM map) for a program $\program$ supported by an invariant map $\inv$ is an $n$-dimensional measurable map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ for $\program$ such that for each configuration $(\loc,\vec{x})$ where $\loc\neq \locterm$ and $\vec{x}\in \inv(\loc)$ the following conditions are satisfied:
 \begin{compactitem}
 	%\item For all $(\loc,\vec{x})\in \locs\times\Rset^{|\pvars|}$ it holds
 	%\item For all  it holds 
 	\item
 	for all $1\leq j \leq n$, $\lem_j(\loc,\vec{x})\geq 0$; and
 	\item 
 	for each generalized transition $\tilde{\tau}$ outgoing from $\loc$ such that $\vec{x}\models\guards(\tilde\tau)$ there 
 	exists $1\leq j 
 	\leq$ n such that
 	\begin{compactitem}
 	\item
 	$\tilde{\tau}$ is $\eps$-ranked by $\lem_j$ from $(\loc,\vec{x})$
 	\item
 	for all $1\leq j'<j$ we have that $\tilde{\tau}$ is unaffected by 
 	$\lem_{j'}$ from $(\loc,\vec{x})$.
 	\end{compactitem}
 \end{compactitem}
If additionally $\lem$ is a linear expression map, then we call it a linear $\eps$-LexRSM map ($\eps$-LinLexRSM).
\end{definition}

@@

\begin{theorem}
\label{thm:lexrsm-programs}
Let $\program$ be a probabilistic program. Assume that there exists an $\eps>0$ 
and an $n$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ for 
$\program$ supported 
by some 
invariant map $\inv$. 
Then $\program$ terminates almost surely.
\end{theorem}
\begin{proof}
Let $\sigma$ be any measurable scheduler and $\vecinit\in\vecinitset$ any 
initial variable valuation in $\program$.
We define an $n$-dimensional stochastic process 
$\{\vec{X}_{i}\}_{i=0}^{\infty} $ on the probability space 
$(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$ such 
that for each 
$i\geq 0$ and $1\leq j 
\leq n$ and each run $\run$ we put $\vecseq{X}{i}{j}(\run) = 
\lem_j(\cfg{\sigma}{i}(\run))$. We claim that $\{\vec{X}_{i}\}_{i=0}^{\infty}$ 
is a strict $n$-dimensional $\eps$-LexRSM for the termination time $\ttime$ of $\program$. Clearly 
the process is real-valued, componentwise non-negative, and adapted to the 
canonical filtration of $\natfilt$. It remains to prove that condition (3) in 
Definition~\ref{def:lexrsm} is satisfied. To this end, for each $i\geq 0$ we 
define an almost-sure partition of the set $\{\run\in \OmegaRun\mid 
\ttime(\run) >i\}$ into sets $L^{i}_1,\dots,L^{i}_n$ by putting $L^i_j$ to be 
the set of all runs $\run$ such that $\ttime(\run)>i$ and for $\run$ the index 
$j$ is the smallest one such that the $(i+1)$-th transition on $\run$ is ranked 
by $\lem_j$ from $\cfg{\sigma}{i}(\run)$. Due to 
definition of an $\eps$-LexRSM map such a $j$ exists for all 
$\run\in\{\ttime>i\}$ and hence we indeed have a partition (so $L^i_{n+1}=\emptyset$ for all $i$). It remains to prove 
that irrespective of the initial choice of $\sigma$ and $\vecinit$ it holds, 
for each $1\leq j 
\leq n$ and $j'<j$, that $\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
\natfilt_i]\leq X_i[j]-\eps $ on $L_j^i$ and 
$\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
\natfilt_i]\leq X_i[j] $ on $L_{j'}^i$. This follows easily from 
the definition of $L^{i}_1,\dots,L^{i}_n$ and from the definition of a 
transition being $\eps$-ranked by $\lem_j$.

>>>>

\begin{definition}[Measurable Maps and Linear Expression Maps]
A 1-dimensional \emph{measurable map} for a \PP{} $\program$ is a  
real-valued function $\lem$ 
assigning to each program location $\loc$ of $\pCFG_P$ a Borel-measurable function $\lem(\loc)$  of program variables, i.e. each $\lem(\loc)$  is a function of type $\Rset^{|\pvars|}\rightarrow \Rset$. As a special case, if all the functions $\lem(\loc)$ are affine, then we call $\lem$ a 1-dimensional \emph{linear expression map (LEM)}. 
Am $n$-dimensional measurable/linear expression map is a vector 
$\overrightarrow{\lem}=(\lem_1,\dots,\lem_n)$ of 1-dimensional 
measurable/linear expression maps. 
\end{definition}

@@

\begin{definition}[Lexicographic Ranking Supermartingale Map]
Let $\eps>0$. An $n$-dimensional \emph{lexicographic $\eps$-ranking 
supermartingale map} ($\eps$-LexRSM map) for a program $\program$ supported by 
an invariant map $\inv$ is an $n$-dimensional measurable map 
$\overrightarrow{\lem}=(\lem_1,\dots,\lem_n)$ for $\program$ such that for each 
configuration $(\loc,\vec{x})$ where $\loc\neq \locterm$ and $\vec{x}\in 
\inv(\loc)$ the following conditions are satisfied:
 \begin{compactitem}
 	%\item For all $(\loc,\vec{x})\in \locs\times\Rset^{|\pvars|}$ it holds
 	%\item For all  it holds 
 	\item
 	for all $1\leq j \leq n$, $\lem_j(\loc,\vec{x})\geq 0$; and
 	\item 
 	for each generalized transition $\tilde{\tau}$ outgoing from $\loc$ such that $\vec{x}\models\guards(\tilde\tau)$ there 
 	exists $1\leq j 
 	\leq$ n such that
 	\begin{compactitem}
 	\item
 	$\tilde{\tau}$ is $\eps$-ranked by $\lem_j$ from $(\loc,\vec{x})$
 	\item
 	for all $1\leq j'<j$ we have that $\tilde{\tau}$ is unaffected by 
 	$\lem_{j'}$ from $(\loc,\vec{x})$.
 	\end{compactitem}
 \end{compactitem}
If additionally $\lem$ is a linear expression map, then we call it a linear $\eps$-LexRSM map ($\eps$-LinLexRSM).
\end{definition}

@@

\begin{theorem}
\label{thm:lexrsm-programs}
Let $\program$ be a probabilistic program. Assume that there exists an $\eps>0$ 
and an $n$-dimensional $\eps$-LexRSM map 
$\overrightarrow{\lem}=(\lem_1,\dots,\lem_n)$ for 
$\program$ supported 
by some 
invariant map $\inv$. 
Then $\program$ terminates almost surely.
\end{theorem}
\begin{proof}
Let $\sigma$ be any measurable scheduler and $\vecinit\in\vecinitset$ any 
initial variable valuation in $\program$.
We define an $n$-dimensional stochastic process 
$\{\vec{X}_{i}\}_{i=0}^{\infty} $ on the probability space 
$(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$ such 
that for each 
$i\geq 0$ and $1\leq j 
\leq n$ and each run $\run$ we put $\vecseq{X}{i}{j}(\run) = 
\lem_j(\cfg{\sigma}{i}(\run))$. We claim that $\{\vec{X}_{i}\}_{i=0}^{\infty}$ 
is a strict $n$-dimensional $\eps$-LexRSM for the termination time $\ttime$ of $\program$. Clearly 
the process is real-valued, componentwise non-negative, and adapted to the 
canonical filtration of $\natfilt$. It remains to prove that condition (3) in 
Definition~\ref{def:lexrsm} is satisfied. To this end, for each $i\geq 0$ we 
define an almost-sure partition of the set $\{\run\in \OmegaRun\mid 
\ttime(\run) >i\}$ into sets $L^{i}_1,\dots,L^{i}_n$ by putting $L^i_j$ to be 
the set of all runs $\run$ such that $\ttime(\run)>i$ and for $\run$ the index 
$j$ is the smallest one such that the $(i+1)$-th transition on $\run$ is ranked 
by $\lem_j$ from $\cfg{\sigma}{i}(\run)$. Due to 
definition of an $\eps$-LexRSM map such a $j$ exists for all 
$\run\in\{\ttime>i\}$ and hence we indeed have a partition (so $L^i_{n+1}=\emptyset$ for all $i$). It remains to prove 
that irrespective of the initial choice of $\sigma$ and $\vecinit$ it holds, 
for each $1\leq j 
\leq n$ and $j'<j$, that $\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
\natfilt_i]\leq X_i[j]-\eps $ on $L_j^i$ and 
$\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
\natfilt_i]\leq X_i[j] $ on $L_{j'}^i$. This follows easily from 
the definition of $L^{i}_1,\dots,L^{i}_n$ and from the definition of a 
transition being $\eps$-ranked by $\lem_j$.

### 2017-07-07 paper/prelims.tex

In this subsection we define the form of probabilistic programs that we 
consider in our analysis. We consider two classes of probabilistic programs: 
general probabilistic programs (\PP{}s ) with arbitrary (measurable) 
expressions and 
their subclass, affine probabilistic programs (\APP s) where all expressions 
are restricted to be affine (see below for a precise definition). The reason 
for this dual view is that our work also has two main points of focus: a 
theoretical one, where we introduce new proof rules that can be used to prove 
properties of general probabilistic programs; and an algorithmic one, where we 
aim to prove properties of probabilistic programs automatically, using the 
aforementioned proof rules. As already testified in the non-probabilistic 
world, programs that contain only affine expressions allow for more efficient 
automation of the analysis and at the same time, due to the presence of 
non-determinism they can be used to form sound abstractions of programs with 
non-linear arithmetic. Hence, we consider general programs when providing our 
theoretical results and \APP s when presenting the automation of our techniques.

@@

\noindent{\em The Syntax of Probabilistic Programs {(\PP s)}.}
We consider the standard syntax for probabilistic programs,
which encompasses basic programming mechanisms such as assignment statement 
(indicated by `:='), while-loop, if-branch. Expressions appear on right-hand sides of assignments, and predicates act as loop guards and conditions in if-then-else statements. We also consider basic probabilistic mechanisms 
such as probabilistic branch (indicated by `prob') and random sampling (e.g. 
$x:=\textbf{sample(}\mathrm{Uniform}[-2,1]\textbf{)}$ assigns to $x$ a random 
number 
uniformly sampled from interval $[-2,1]$). We also allow constructs for 
(demonic) non-determinism, in particular 
non-deterministic branching indicated by `\textbf{if }$\star$ \textbf{then...}' construct and non-deterministic assignment.  
Variables (or identifiers) of a probabilistic program are of \emph{real} type, i.e., 
values of the variables are real numbers. 
We also assume that assume that each \PP{} $\program$ is preceded by a preamble 
specifying possible initial values of program variables: the preamble consists 
of a single predicate characterizing possible initial valuations.

@@

\emph{Distributions.} We assume the standard definition of a probability 
distribution specified by a cumulative distribution 
function~\cite{Billingsley:book}. We denote by $\mathcal{D}$ be a set of 
probability distributions on 
real numbers, both discrete and continuous.

@@

We assume that each location has at least one outgoing transition.
Also, for every deterministic location $\ell$ we assume the following: if
$\tau_1,\dots,\tau_k$ are all transitions outgoing from $\ell$, then $G(\tau_1)
\vee \dots \vee G(\tau_k) \equiv \mathit{true}$ and $G(\tau_i) \wedge G(\tau_j)
\equiv \mathit{false}$ for each $1\leq i < j \leq k$. For each 
distribution $d$ appearing in the 
pCFG we assume the following features are known: expected value $\expv[d]$ of 
$d$ and a set $SP_d$ containing the \emph{support} of $d$. Tthe support is the  
smallest 
closed set of real numbers whose complement has probability zero 
under $d$\footnote{In particular, a support of a \emph{discrete} probability 
	distribution $d$ is simply the at most countable set of all points on a 
	real 
	line that have positive probability under $d$.}) Finally, we assume that 
	each 
assignment location has at most (and thus exactly) one outgoing transition. The 
translation from probabilistic programs to the corresponding pCFG is 
standard~\cite{CFNH16:prob-termination}, and the details are presented 
in~\AppendixMaterial.
%Finally, we assume that for each 
%transition $(\loc,j,u,\loc')$ such that $u$ is a set the location $\loc$ is 
%deterministic. This is just a technical assumption yielding no loss of 
%generality, and it somewhat simplifies notation.
%And we assume that each coordinate of $\mathcal{D}$ represents an integrable
%random variable (i.e., the expected value of the absolute value of the random
%variable exists).
\end{definition}

>>>>

In this subsection we define the form of probabilistic programs that we 
consider in our analysis. We consider two classes of probabilistic programs: 
general probabilistic programs (\PP{}s) with arbitrary (measurable) 
expressions and 
their subclass, affine probabilistic programs (\APP s) where all expressions 
are restricted to be affine (see below for a precise definition). The reason 
for this dual view is that our work also has two main points of focus: a 
theoretical one, where we introduce new concepts that can be used to prove 
properties of general probabilistic programs; and an algorithmic one, where we 
aim to prove properties of probabilistic programs automatically, using the 
aforementioned concepts. As already testified in the non-probabilistic 
world, programs that contain only affine expressions allow for more efficient 
automation of the analysis and at the same time, due to the presence of 
non-determinism they can be used to form sound abstractions of programs with 
non-linear arithmetic. Hence, we consider general programs when providing our 
theoretical results and \APP s when presenting the automation of our techniques.

@@

\noindent{\em The Syntax of Probabilistic Programs {(\PP s)}.}
We consider the standard syntax for probabilistic programs,
which encompasses basic programming mechanisms such as assignment statement 
(indicated by `:='), while-loop, if-branch. Expressions appear on right-hand sides of assignments, and predicates act as loop guards and conditions in if-then-else statements. We also consider basic probabilistic mechanisms 
such as probabilistic branch (indicated by `prob') and random sampling (e.g. 
$x:=\textbf{sample(}\mathrm{Uniform}[-2,1]\textbf{)}$ assigns to $x$ a random 
number 
uniformly sampled from interval $[-2,1]$). We also allow constructs for 
(demonic) non-determinism, in particular 
non-deterministic branching indicated by `\textbf{if }$\star$ \textbf{then...}' construct and non-deterministic assignment.  
Variables (or identifiers) of a probabilistic program are of \emph{real} type, i.e., 
values of the variables are real numbers. 
We also assume that each \PP{} $\program$ is preceded by a preamble 
specifying possible initial values of program variables: the preamble consists 
of a single predicate characterizing possible initial valuations.

@@

\emph{Distributions.} We assume the standard definition of a probability 
distribution specified by a cumulative distribution 
function~\cite{Billingsley:book}. We denote by $\mathcal{D}$ the set of 
probability distributions on 
real numbers, both discrete and continuous.

@@

We assume that each location has at least one outgoing transition.
Also, for every deterministic location $\ell$ we assume the following: if
$\tau_1,\dots,\tau_k$ are all transitions outgoing from $\ell$, then $G(\tau_1)
\vee \dots \vee G(\tau_k) \equiv \mathit{true}$ and $G(\tau_i) \wedge G(\tau_j)
\equiv \mathit{false}$ for each $1\leq i < j \leq k$. For each 
distribution $d$ appearing in the 
pCFG we assume the following features are known: expected value $\expv[d]$ of 
$d$ and a set $SP_d$ containing the \emph{support} of $d$. The support is the  
smallest 
closed set of real numbers whose complement has probability zero 
under $d$.\footnote{In particular, a support of a \emph{discrete} probability 
	distribution $d$ is simply the at most countable set of all points on a 
	real 
	line that have positive probability under $d$.} Finally, we assume that 
	each 
assignment location has at most (and thus exactly) one outgoing transition. The 
translation from probabilistic programs to the corresponding pCFG is 
standard~\cite{CFNH16:prob-termination}, and the details are presented 
in~\AppendixMaterial.
%Finally, we assume that for each 
%transition $(\loc,j,u,\loc')$ such that $u$ is a set the location $\loc$ is 
%deterministic. This is just a technical assumption yielding no loss of 
%generality, and it somewhat simplifies notation.
%And we assume that each coordinate of $\mathcal{D}$ represents an integrable
%random variable (i.e., the expected value of the absolute value of the random
%variable exists).
\end{definition}

### 2017-07-07 paper/synthesis.tex

Both soundness and relative completeness of Algorithm~\ref{algo:linlexrsm} 
can be proved by generalization of arguments presented 
in~\cite{ADFG10:lexicographic}. 
\begin{theorem}
\label{thm:algo}
\label{THM:ALGO}
	Suppose that Algorithm~\ref{algo:linlexrsm} is run on an \APP{} $\program$ 
	together with linear invariant map $\inv$. If the algorithm returns a 
	$d$-dimensional LEM $\vec{\lem}=(\lem_1,\dots,\lem_d)$, then $\vec{\lem}$ 
	is a $1$-LinLexRSM map for $\program$ supported by $\inv$. Conversely, if 
	the algorithm returns ``No LinLexRSM'', then for any $d'\in \Nset$ and 
	$\eps>0$ there is no $d'$-dimensional $\eps$-LinLexRSM for $\program$  
	supported by $\inv$. If guards of all conditionals and loops in $\APP{}$ 
	are linear assertion, then the
	algorithm runs in time polynomial in size of 
	$\program$ and $\inv$.
\end{theorem}
\begin{proof}[Proof (Key Ideas)]
For soundness, let $U_i$ and $U_{i}$ denote the content of $U$ just before the $i$-th iteration of the while loop. Then each $\tilde{\tau}\in U_{i}\setminus U_{i+1}$ is $1$-ranked by $\lem_i$, for each $1\leq i \leq d$. Since $U_{d}=\emptyset$, each generalized transition of $\pCFG_{\program}$ is 1-ranked by some component of $\vec{\lem}$. Non-negativity of each $\lem_i$ is ensured directly by $\lp_{U_i}$. Hence, it remains to show that each $\tilde{\tau}\in U_i$ is unaffected by $\lem_1,\dots,\lem_{i-1}$. But this follows from $U_i \subseteq U_{i-1}\subseteq \cdots\subseteq U_1$ and from the fact that each gen. transition in $U_j$ is unaffected by $\lem_j$.

>>>>

Both soundness and relative completeness of Algorithm~\ref{algo:linlexrsm} 
can be proved by generalization of arguments presented 
in~\cite{ADFG10:lexicographic}. 
\begin{theorem}
\label{thm:algo}
\label{THM:ALGO}
	Suppose that Algorithm~\ref{algo:linlexrsm} is run on an \APP{} $\program$ 
	together with linear invariant map $\inv$. If the algorithm returns a 
	$d$-dimensional LEM $\overrightarrow{\lem}=(\lem_1,\dots,\lem_d)$, then 
	$\overrightarrow{\lem}$ 
	is a $1$-LinLexRSM map for $\program$ supported by $\inv$. Conversely, if 
	the algorithm returns ``No LinLexRSM'', then for any $d'\in \Nset$ and 
	$\eps>0$ there is no $d'$-dimensional $\eps$-LinLexRSM for $\program$  
	supported by $\inv$. If guards of all conditionals and loops in $\APP{}$ 
	are linear assertion, then the
	algorithm runs in time polynomial in size of 
	$\program$ and $\inv$.
\end{theorem}
\begin{proof}[Proof (Key Ideas)]
For soundness, let $U_i$ and $U_{i}$ denote the content of $U$ just before the 
$i$-th iteration of the while loop. Then each $\tilde{\tau}\in U_{i}\setminus 
U_{i+1}$ is $1$-ranked by $\lem_i$, for each $1\leq i \leq d$. Since 
$U_{d}=\emptyset$, each generalized transition of $\pCFG_{\program}$ is 
1-ranked by some component of $\overrightarrow{\lem}$. Non-negativity of each 
$\lem_i$ is ensured directly by $\lp_{U_i}$. Hence, it remains to show that 
each $\tilde{\tau}\in U_i$ is unaffected by $\lem_1,\dots,\lem_{i-1}$. But this 
follows from $U_i \subseteq U_{i-1}\subseteq \cdots\subseteq U_1$ and from the 
fact that each gen. transition in $U_j$ is unaffected by $\lem_j$.
